### 2017-12-12 submission/manuscript.Rmd
complete imporvements to disussion -- reduce causation talk and improve flow

Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two conditions. Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma. This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community contributing to the formation of an adenoma. Additionally, our data show that by using the whole community, our models perform better than when only the CRC-associated genera are included. CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions. These observations suggest that small localized changes within the community on tissue may be occuring that are important in early disease progression of CRC and that this process may not directly involve CRC-associated genera.

The data presented herein supports the driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], when applied to carcinoma but not necessarily adenoma. The central idea of the model is that a single bacterium initiates an environment in which other non-resident microbes may then be able to colonize, creating a vicious cycle that is conducive for CRC. Both the drastically increased RR of CRC-associated genera versus $\alpha$-diversity metrics for carcinoma and increasing RR with more CRC-associated generea positivity are highly supportive of this model. However, the initial establishment of the driver within the system appears to be dependent on the current community. This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases [Figure 4 & 5]. Conversely, using the present data, it is less likely that adenoma development fits this model. The changes that occur at this timepoint are small and possibly focal to the adenoma itself. The stepwise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change, to a change in the global community. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. One potential hypothesis from these observations is that at early stages of the diease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6]. Changes in *Bacteroides*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma models [Figure 6]. This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. These observations would also suggest that initial changes in the resident community, specifically to *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma. Based on these observations, it is possible to hypothesize that the early changes in the community may give rise to initial polyp formation via interactions with the host and not necessarily via interactions with CRC-assciated genera. These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the microbiota may drive the progression of disease severity. For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 important variables across studies. Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F]. The majority of the results seem to be study specific with many top 10 taxa being present only in a single study. One other potentially worrying sign is the presence of *Propionibacterium* within the top 10 for the genera and OTU models and could be a marker of contamination. The low biomass of these samples coupled with potential contamination might be a possible reason why the tissue results seem to be more sporadic then the stool results.

Although there are still questions that need to be answered regarding the microbiota and carcinoma, a clearer framework for their relationship is beginning to develop as to how this occurs. From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 7]. Conversely, the associations between the microbiota and adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls. This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out how the microbiota contributes to adenoma formation.

>>>>

Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two. Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma. This information suggests that these specific genera are strongly associated with carcinoma but may not be associated with adenoma. Although CRC-associated genera are asscoiated with carcinomas, our data show that by using the whole community, classification models perform better than when only the CRC-associated genera are included. CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions. These observations suggest that small localized changes within the community on tissue may be occuring, that are important in early disease progression of CRC, and that this process may not directly involve CRC-associated genera.

The data presented herein supports the driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], when applied to carcinoma but not necessarily adenoma. The central idea of the model is that a single bacterium initiates an environment in which other non-resident microbes may then be able to colonize, creating a vicious cycle that is conducive for CRC. Both the drastically increased RR of CRC-associated genera versus $\alpha$-diversity metrics for carcinoma and increasing RR with more CRC-associated genera positivity are highly supportive of this model. However, the initial establishment of the driver within the system appears to be dependent on the current community. This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases [Figure 4 & 5]. Conversely, using the present data, it is less likely that adenoma development fits this model. The changes that occur at this timepoint are small and possibly focal to the adenoma itself. The stepwise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change, to a change in the global community. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6]. Changes in *Bacteroides*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma models [Figure 6]. This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. These observations would also suggest that initial changes in the resident community, specifically to *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma. Based on these observations, it is possible to hypothesize that the early changes in the community may give rise to initial polyp formation via interactions with the host and not necessarily via interactions with CRC-assciated genera. These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the microbiota may be associated with disease severity. For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 important variables across studies. Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F]. The majority of the results seem to be study specific with many top 10 taxa being present only in a single study. One other potentially worrying sign is the presence of *Propionibacterium* within the top 10 for the genera and OTU models and could be a marker of contamination. The low biomass of these samples coupled with potential contamination might be a possible reason why the tissue results seem to be more sporadic then the stool results.

The associations between the microbiota and adenoma are less clear in part because many studies may not be powered effectively to observe the small changes reported here. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls. This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out how the microbiota contributes to adenoma formation. In contrast to adenoma, our observations suggest that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 7].

### 2017-12-12 submission/manuscript.tex
complete imporvements to disussion -- reduce causation talk and improve flow

Our study identifies clear differences in diversity both at the
community level and within individual genera that are present in
individuals with CRC versus those without the disease {[}Figure 1-3{]}.
Although there was a step-wise decrease in diversity from control to
adenoma to carcinoma, this did not translate into large effect sizes for
the relative risk of either of these two conditions. Even though
CRC-associated genera increase the RR of carcinoma, they do not
consistently increase the relative risk of adenoma. This information
suggests that these specific genera are important in carcinoma genesis
but may not be the primary members of the microbial community
contributing to the formation of an adenoma. Additionally, our data show
that by using the whole community, our models perform better than when
only the CRC-associated genera are included. CRC-associated genera are
clearly important to carcinoma pathogenesis but accounting for the
community in which these microbes exist can drastically increase the
ability of models to make predictions. These observations suggest that
small localized changes within the community on tissue may be occuring
that are important in early disease progression of CRC and that this
process may not directly involve CRC-associated genera.

The data presented herein supports the driver-passenger model of the
microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
carcinoma but not necessarily adenoma. The central idea of the model is
that a single bacterium initiates an environment in which other
non-resident microbes may then be able to colonize, creating a vicious
cycle that is conducive for CRC. Both the drastically increased RR of
CRC-associated genera versus \(\alpha\)-diversity metrics for carcinoma
and increasing RR with more CRC-associated generea positivity are highly
supportive of this model. However, the initial establishment of the
driver within the system appears to be dependent on the current
community. This is supported by our finding that when adding the
community context to our models in addition to the CRC-associated
genera, the model AUC increases {[}Figure 4 \& 5{]}. Conversely, using
the present data, it is less likely that adenoma development fits this
model. The changes that occur at this timepoint are small and possibly
focal to the adenoma itself. The stepwise decrease in diversity suggests
that the adenoma community is not normal but has changed subtly
{[}Figure 1{]}. Although there appears to be localized changes that do
depend on the driver-passenger model, as supported by an increased RR
for one, two, and four positive CRC-associated genera in tissue
{[}Figure 3C{]}, there may be other processes at play that ultimately
exacerbate the condition from a subtle localized change, to a change in
the global community. The poor performance of the Random Forest models
for classifying adenoma based only on the microbiota would suggest that
this is the case. One potential hypothesis from these observations is
that at early stages of the diease, how the host interacts with these
subtle changes is what ultimately leads to a thoroughly dysfunctional
community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be disciminative across the different studies for
both adenoma and carcinoma models {[}Figure 6{]}. This data would
suggest that whether the non-resident bacterium is \emph{Fusobacteria}
or \emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. These observations would
also suggest that initial changes in the resident community,
specifically to \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia}, carry on from adenoma to carcinoma. Based on these
observations, it is possible to hypothesize that the early changes in
the community may give rise to initial polyp formation via interactions
with the host and not necessarily via interactions with CRC-assciated
genera. These changes then create new niches in which any one of the
CRC-associated genera could gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma via a driver-passenger type mechanism.

The tissue studies did not provide a clearer understanding of how the
microbiota may drive the progression of disease severity. For the OTU
models both the unmatched and matched {[}Figure S7E \& F{]} tissue
samples had some concordance with the stool data, with resident bacteria
being the most prevalent in the top 10 important variables across
studies. Unlike in stool, \emph{Fusobacterium} was the only
CRC-associated bacteria consistently present in the top 10 of the CRC
models {[}Figure S7B-C \& E-F{]}. The majority of the results seem to be
study specific with many top 10 taxa being present only in a single
study. One other potentially worrying sign is the presence of
\emph{Propionibacterium} within the top 10 for the genera and OTU models
and could be a marker of contamination. The low biomass of these samples
coupled with potential contamination might be a possible reason why the
tissue results seem to be more sporadic then the stool results.

Although there are still questions that need to be answered regarding
the microbiota and carcinoma, a clearer framework for their relationship
is beginning to develop as to how this occurs. From our observations
many changes in carcinoma could easily result in effect sizes that are
30\% or more between the case and control and most studies analyzed have
sufficient power to detect these types of changes {[}Figure 7{]}.
Conversely, the associations between the microbiota and adenoma is less
clear and part of the reason this may be is because many studies are not
powered effectively to observe the small changes reported here. None of
the studies analyzed were properly powered to detect a 10\% or lower
change between case and controls. This small effect size range may well
be the scope in which differences consistently occur in adenoma due to
the subtle changes in community that occur between control and adenoma.
Future studies investigating adenoma and the microbiota need to take
these factors into consideration if we are to work out how the
microbiota contributes to adenoma formation.

>>>>

Our study identifies clear differences in diversity both at the
community level and within individual genera that are present in
individuals with CRC versus those without the disease {[}Figure 1-3{]}.
Although there was a step-wise decrease in diversity from control to
adenoma to carcinoma, this did not translate into large effect sizes for
the relative risk of either of these two. Even though CRC-associated
genera increase the RR of carcinoma, they do not consistently increase
the relative risk of adenoma. This information suggests that these
specific genera are strongly associated with carcinoma but may not be
associated with adenoma. Although CRC-associated genera are asscoiated
with carcinomas, our data show that by using the whole community,
classification models perform better than when only the CRC-associated
genera are included. CRC-associated genera are clearly important to
carcinoma pathogenesis but accounting for the community in which these
microbes exist can drastically increase the ability of models to make
predictions. These observations suggest that small localized changes
within the community on tissue may be occuring, that are important in
early disease progression of CRC, and that this process may not directly
involve CRC-associated genera.

The data presented herein supports the driver-passenger model of the
microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
carcinoma but not necessarily adenoma. The central idea of the model is
that a single bacterium initiates an environment in which other
non-resident microbes may then be able to colonize, creating a vicious
cycle that is conducive for CRC. Both the drastically increased RR of
CRC-associated genera versus \(\alpha\)-diversity metrics for carcinoma
and increasing RR with more CRC-associated genera positivity are highly
supportive of this model. However, the initial establishment of the
driver within the system appears to be dependent on the current
community. This is supported by our finding that when adding the
community context to our models in addition to the CRC-associated
genera, the model AUC increases {[}Figure 4 \& 5{]}. Conversely, using
the present data, it is less likely that adenoma development fits this
model. The changes that occur at this timepoint are small and possibly
focal to the adenoma itself. The stepwise decrease in diversity suggests
that the adenoma community is not normal but has changed subtly
{[}Figure 1{]}. Although there appears to be localized changes that do
depend on the driver-passenger model, as supported by an increased RR
for one, two, and four positive CRC-associated genera in tissue
{[}Figure 3C{]}, there may be other processes at play that ultimately
exacerbate the condition from a subtle localized change, to a change in
the global community. The poor performance of the Random Forest models
for classifying adenoma based only on the microbiota would suggest that
this is the case. One potential hypothesis from these observations is
that at early stages of the disease, how the host interacts with these
subtle changes is what ultimately leads to a thoroughly dysfunctional
community that is supportive of CRC genesis.

Within stool, common resident microbes were most consistently present in
the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be disciminative across the different studies for
both adenoma and carcinoma models {[}Figure 6{]}. This data would
suggest that whether the non-resident bacterium is \emph{Fusobacteria}
or \emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. These observations would
also suggest that initial changes in the resident community,
specifically to \emph{Bacteroides}, \emph{Ruminococcaceae},
\emph{Ruminococcus}, and \emph{Roseburia}, carry on from adenoma to
carcinoma. Based on these observations, it is possible to hypothesize
that the early changes in the community may give rise to initial polyp
formation via interactions with the host and not necessarily via
interactions with CRC-assciated genera. These changes then create new
niches in which any one of the CRC-associated genera could gain a
foothold, exacerbating the initial changes in community and facilitating
the transition from adenoma to carcinoma via a driver-passenger type
mechanism.

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with disease severity. For the OTU models
both the unmatched and matched {[}Figure S7E \& F{]} tissue samples had
some concordance with the stool data, with resident bacteria being the
most prevalent in the top 10 important variables across studies. Unlike
in stool, \emph{Fusobacterium} was the only CRC-associated bacteria
consistently present in the top 10 of the CRC models {[}Figure S7B-C \&
E-F{]}. The majority of the results seem to be study specific with many
top 10 taxa being present only in a single study. One other potentially
worrying sign is the presence of \emph{Propionibacterium} within the top
10 for the genera and OTU models and could be a marker of contamination.
The low biomass of these samples coupled with potential contamination
might be a possible reason why the tissue results seem to be more
sporadic then the stool results.

The associations between the microbiota and adenoma are less clear in
part because many studies may not be powered effectively to observe the
small changes reported here. None of the studies analyzed were properly
powered to detect a 10\% or lower change between case and controls. This
small effect size range may well be the scope in which differences
consistently occur in adenoma due to the subtle changes in community
that occur between control and adenoma. Future studies investigating
adenoma and the microbiota need to take these factors into consideration
if we are to work out how the microbiota contributes to adenoma
formation. In contrast to adenoma, our observations suggest that many
changes in carcinoma could easily result in effect sizes that are 30\%
or more between the case and control and most studies analyzed have
sufficient power to detect these types of changes {[}Figure 7{]}.
